---
title: "Case study of a Cyclistic (Divvy) bike share data"
author: "Shibin Shahid K V"
date: "`r Sys.Date()`"
output: html_document
---
 
This is a case study done as part of Google Data Analytics Capstone: Case study of a Cyclistic (Divvy) bike share data.

Installing the required packages
```{r message=FALSE, warning=FALSE}
install.packages("tidyverse")  # tidyverse for data import and wrangling
install.packages("lubridate")  # lubridate for date functions
install.packages("ggplot2")    # ggplot for visualization
install.packages("readxl")     # readxl for uploading excel files
install.packages("janitor")    # for cleaning
```

Load the packages
```{r message=FALSE, warning=FALSE}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)    #helps visualize data
library(readxl)     #helps to read excel file
library(janitor)
```

=====================
## STEP 1: COLLECT DATA
=====================

The source for downloading the data is from the [link](https://divvy-tripdata.s3.amazonaws.com/index.html).
I have done the analysis from October 2021 to September 2022.
Upload Divvy datasets (xlsx files) here
Here I have added columns for ride_length and day_of_week from excel sheet itself
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
tripdata_2021_oct <- read_excel('202110-divvy-tripdata.xlsx')
tripdata_2021_nov <- read_excel('202111-divvy-tripdata.xlsx')
tripdata_2021_dec <- read_excel('202112-divvy-tripdata.xlsx')
tripdata_2022_jan <- read_excel('202201-divvy-tripdata.xlsx')
tripdata_2022_feb <- read_excel('202202-divvy-tripdata.xlsx')
tripdata_2022_mar <- read_excel('202203-divvy-tripdata.xlsx')
tripdata_2022_apr <- read_excel('202204-divvy-tripdata.xlsx')
tripdata_2022_may <- read_excel('202205-divvy-tripdata.xlsx')
tripdata_2022_jun <- read_excel('202206-divvy-tripdata.xlsx')
tripdata_2022_jul <- read_excel('202207-divvy-tripdata.xlsx')
tripdata_2022_aug <- read_excel('202208-divvy-tripdata.xlsx')
tripdata_2022_sep <- read_excel('202209-divvy-tripdata.xlsx')
```

====================================================
## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
====================================================
Compare column names each of the files
checking the datatypes are consistent in all sheets

```{r message=FALSE, warning=FALSE}
str(tripdata_2021_oct)
str(tripdata_2021_nov)
str(tripdata_2021_dec)
str(tripdata_2022_jan)
str(tripdata_2022_feb)
str(tripdata_2022_mar)
str(tripdata_2022_apr)
str(tripdata_2022_may)
str(tripdata_2022_jun)
str(tripdata_2022_jul)
str(tripdata_2022_aug)
str(tripdata_2022_sep) # end_station_id was found to be in num data type. so it has to be converted to chr
```

converting end_station_id of sep_2022 to chr
```{r message=FALSE, warning=FALSE}
tripdata_2022_sep <- mutate(tripdata_2022_sep,
  end_station_id = as.character(end_station_id))
```

Combining the 12 months data into 1 dataframe

```{r message=FALSE, warning=FALSE}
tripdata_combined <- bind_rows(
                              tripdata_2021_oct,
                              tripdata_2021_nov,
                              tripdata_2021_dec,
                              tripdata_2022_jan,
                              tripdata_2022_feb,
                              tripdata_2022_mar,
                              tripdata_2022_apr,
                              tripdata_2022_may,
                              tripdata_2022_jun,
                              tripdata_2022_jul,
                              tripdata_2022_aug,
                              tripdata_2022_sep
                              )
```

======================================================
## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
======================================================
Inspect the new table that has been created

```{r message=FALSE, warning=FALSE}
colnames(tripdata_combined) #List of column names
nrow(tripdata_combined)     #How many rows are in data frame?
dim(tripdata_combined)      #Dimensions of the data frame?
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
head(tripdata_combined)     #See the first 6 rows of data frame.  Also tail(all_trips)
str(tripdata_combined)      #See list of columns and data types (numeric, character, etc)
summary(tripdata_combined)  #Statistical summary of data. Mainly for numerics
```

Finding the ride_length and converting into seconds

```{r message=FALSE, warning=FALSE}
tripdata_combined$ride_length <- difftime(
                                          tripdata_combined$ended_at, 
                                          tripdata_combined$started_at,
                                          units = "secs"
                                          )
```

converting ride_length data type from 'difftime' num to num
```{r message=FALSE, warning=FALSE}
tripdata_combined$ride_length <- as.numeric(tripdata_combined$ride_length)
```

Add columns that list the date, month, day, and year of each ride
This will allow us to aggregate ride data for each month, day, or year .before completing these operations we could only aggregate at the ride level
```{r message=FALSE, warning=FALSE}
tripdata_combined$date <- as.Date(tripdata_combined$started_at)               #The default format is yyyy-mm-dd
tripdata_combined$month <- format(as.Date(tripdata_combined$date), "%m")
tripdata_combined$day <- format(as.Date(tripdata_combined$date), "%d")
tripdata_combined$year <- format(as.Date(tripdata_combined$date), "%Y")
tripdata_combined$day_of_week <- format(as.Date(tripdata_combined$date), "%A")
```

Remove "bad" data
The dataframe includes a few entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
We will create a new version of the dataframe (all_data) since data is being removed
```{r message=FALSE, warning=FALSE}
all_trips <- tripdata_combined %>% filter(ride_length>=0) %>% filter(start_station_name != "HQ QR")
```

## STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
=====================================
Descriptive analysis on ride_length (all figures in seconds)
```{r message=FALSE, warning=FALSE}
mean(all_trips$ride_length) #straight average (total ride length / rides)
median(all_trips$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips$ride_length) #longest ride
min(all_trips$ride_length) #shortest ride
```

Compare members and casual users
```{r echo=TRUE, message=FALSE, warning=FALSE}
aggregate(all_trips$ride_length ~ all_trips$member_casual, FUN = mean)
aggregate(all_trips$ride_length ~ all_trips$member_casual, FUN = median)
aggregate(all_trips$ride_length ~ all_trips$member_casual, FUN = max)
aggregate(all_trips$ride_length ~ all_trips$member_casual, FUN = min)
```

See the average ride time by each day for members vs casual users
```{r echo=TRUE, message=FALSE, warning=FALSE}
aggregate(all_trips$ride_length ~ all_trips$member_casual + all_trips$day_of_week, FUN = mean)
```

Notice that the days of the week are out of order. Let's fix that.
```{r message=FALSE, warning=FALSE}
all_trips$day_of_week <- ordered(all_trips$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Now, let's run the average ride time by each day for members vs casual users
```{r echo=TRUE, message=FALSE, warning=FALSE}
aggregate(all_trips$ride_length ~ all_trips$member_casual + all_trips$day_of_week, FUN = mean)
```

analyze ridership data by type and weekday
```{r echo=TRUE, message=FALSE, warning=FALSE}
all_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  # creates weekday field using wday()
  group_by(member_casual, weekday) %>%                  # groups by usertype and weekday
  summarise(number_of_rides = n()						          	# calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 	# calculates the average duration
  arrange(member_casual, weekday)
```

Let's visualize the number of rides by rider type
```{r echo=TRUE, message=FALSE, warning=FALSE}
all_trips %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

Let's create a visualization for average duration
```{r echo=TRUE, message=FALSE, warning=FALSE}
all_trips%>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

